What to Expect with Life Expectancy?

### 2024-11-27 paper/paper.qmd
Add files via upload

This study aims to address these gaps by developing a linear model to analyze the impact of a nation’s status, healthcare factors, and economic indicators on life expectancy. The response variable is Life Expectancy, while the predictors include Status (Developed vs. Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI. The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.

@@

Together, these datasets enable an in-depth exploration of the relationships between life expectancy and various predictors, such as economic indicators, healthcare access, and immunization coverage. Key variables include Life Expectancy, Status (Developed vs. Developing), Diphtheria Immunization Coverage, Government Health Expenditure as percentage of Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI.

@@

**Body Mass Index (`BMI`)**: This variable captures the average body mass index of the population, calculated as weight in kilograms divided by the square of height in meters. It is an aggregate measure derived from population health surveys and recorded as a numeric value.

@@

2. **Body Mass Index (`BMI`)**: This variable represents the average Body Mass Index of the entire population in a country. BMI is a widely used health indicator that reflects the prevalence of obesity or underweight populations. Extreme values of BMI can negatively impact life expectancy by contributing to chronic diseases such as diabetes and cardiovascular conditions. The BMI condition is depicted in @fig-bmi.

3. **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization. Immunization coverage(@fig-diphtheria) is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.

4. **Schooling (`Schooling`)**: The number of years of schooling in a country is used as a proxy for education levels. Higher education levels typically correlate with better health outcomes and, consequently, higher life expectancy, as educated individuals are more likely to engage in healthier behaviors and have access to better healthcare. The Educated condition is shown in @fig-schooling.

5. **Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor). Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.

6. **Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing). Nurses play a crucial role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.

@@

```{r}
#| label: fig-bmi
#| fig-cap: "Density of BMI by Status, showing a higher concentration of BMI in developed regions around 24, suggesting a potential focus on obesity if the bin width were wider. The distribution for developing regions is fairly evenly distributed, with less concentration in any specific BMI range."
#| echo: false
#| warning: false
#| message: false

data_filtered <- data %>%
  filter(BMI >= 18.5 & BMI <= 24.9)
ggplot(data_filtered, aes(x = BMI, fill = Status)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "BMI Condition by Status (Normal Range)",
    x = "BMI",
    y = "Density"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("royalblue", "red")) +  # Customize colors for Developed and Developing
  facet_wrap(~Status)

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...
represent socioeconomic predictors such as BMI, Diphtheria, Schooling, TotalExpenditure, pct_doctor, and pct_nursing.

```{=tex}
\begin{align*} 
\text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \epsilon_i
\end{align*}

\begin{align*} 
\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{BMI} + \beta_2 \text{Diphtheria Immunization} + \\
&\ \beta_3 \text{Schooling} + \beta_4 \text{Number of Doctors per 10000} + \\
&\ \beta_5 \text{Number of Nurses per 10000} + \epsilon_i
\end{align*}
```

@@

The models were built using linear regression (LM), which is appropriate for predicting life expectancy as a continuous variable. This method captures the relationship between predictors and the response variable with interpretable coefficients. For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries. In the Developing model, we added BMI, Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations. Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.

@@

**BMI**: Coefficient = 0.040, Standard Error = 0.012

@@

In the Developing model, `Schooling` again has the most significant impact, and `#Doctors per 10000` is also a substantial predictor. This highlights the importance of healthcare workforce availability in improving life expectancy in developing countries, unlike the developed context where Schooling was more prominent. The negative relationship of `Total Expenditure` suggests that in developing countries, spending in certain healthcare sectors could be misallocated or inefficient.
```{r}
#| label: tbl-keyestimates
#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like BMI, Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate. Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

model_list <- list("Model for Developed" = lm_developed, "Model for Developing" = lm_developing)
modelsummary(model_list, 
             output = "kableExtra", 
             statistic = "std.error", 
             coef_map = c("BMI" = "BMI", 
                          "Diphtheria" = "Diphtheria Immunization",
                          "Schooling" = "Schooling",
                          "TotalExpenditure" = "Health Expenditure",
                          "pct_doctor" = "Number of Medical Doctors per 10000",
                          "pct_nursing" = "Number of Nurses per 10000")) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| echo: false
#| label: tbl-modelresults
#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 77.23 years, while in developing countries, it is slightly lower at 72.1 years. This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries. These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and advanced medical technology, contributing to longer life expectancies. "
#| eval: true
#| warning: false
#| message: false

>>>>

This study aims to address these gaps by developing a linear model to analyze the impact of a nation’s status, healthcare factors, and economic indicators on life expectancy. The response variable is Life Expectancy, while the predictors include Status (Developed vs. Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling. The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.

@@

Together, these datasets enable an in-depth exploration of the relationships between life expectancy and various predictors, such as economic indicators, healthcare access, and immunization coverage. Key variables include Life Expectancy, Status (Developed vs. Developing), Diphtheria Immunization Coverage, Government Health Expenditure as percentage of Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling.

@@

@@

2. **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization. Immunization coverage(@fig-diphtheria) is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.

3. **Schooling (`Schooling`)**: The number of years of schooling in a country is used as a proxy for education levels. Higher education levels typically correlate with better health outcomes and, consequently, higher life expectancy, as educated individuals are more likely to engage in healthier behaviors and have access to better healthcare. The Educated condition is shown in @fig-schooling.

4. **Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor). Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.

5. **Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing). Nurses play a crucial role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.

@@

@@

The goal of this analysis is to construct a linear model to predict life expectancy (`LifeExpectancy`), which represents the average number of years an individual is expected to live. This model incorporates several predictor variables that capture economic and social factors influencing life expectancy.

@@

The models assume a linear relationship between the predictors and life expectancy, where each predictor contributes independently or interactively to the variation in the outcome variable. Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria, Schooling, TotalExpenditure, pct_doctor, pct_nursing, and interaction term.

The linear regression model for developed and developing regions are specified as follows:

```{=tex}

\begin{align*} 
\text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
&\ \beta_5 \text{Number of Nurses per 10000} + \\
&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
\end{align*}

\begin{align*} 
\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
&\ \beta_5 \text{Number of Nurses per 10000} + \\
&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
\end{align*}

Where:

-   `LifeExpectancy`: Represents the average number of years a person is expected to live in a specific country, based on current mortality trends.

-   `Diphtheria Immunization`: The percentage of children who received the diphtheria-tetanus-pertussis (DTP3) vaccine, serving as a proxy for the quality and reach of immunization programs.

-   `Schooling`: Average years of schooling for individuals aged 25 and above, reflecting the level of education in the population.

-   `Health Expenditure`: The percentage of government expenditure allocated to healthcare, indicating the priority given to public health.

-   `Number of Doctors per 10,000`: The availability of medical doctors per 10,000 people, capturing the accessibility of professional healthcare services.

-   `Number of Nurses per 10,000`: The availability of nursing staff per 10,000 people, further representing healthcare capacity.

-   Interaction Term: The combined effect of health expenditure, doctors, and nurses on life expectancy, considering how these variables may amplify or mitigate each other’s impact.

-   epsilon is the error term, assumed to be normally distributed with a mean of 0 and constant variance $ϵ_i​∼N(0,\sigma^2)$

@@

The models were built using linear regression (LM), which is appropriate for predicting life expectancy as a continuous variable. This method captures the relationship between predictors and the response variable with interpretable coefficients. For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries. In the Developing model, we added Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations. Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.

@@

@@

In the Developing model, `Schooling` again has the most significant impact, and `#Doctors per 10000` is also a substantial predictor. This highlights the importance of healthcare workforce availability in improving life expectancy in developing countries, unlike the developed context where Schooling was more prominent. The negative relationship of `Total Expenditure` suggests that in developing countries, spending in certain healthcare sectors could be misallocated or inefficient.
```{r}
#| label: tbl-keyestimates
#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate. Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

model_list <- list("Model for Developed" = lm_developed, "Model for Developing" = lm_developing)
modelsummary(model_list, 
             output = "kableExtra", 
             statistic = "std.error", 
             coef_map = c("Diphtheria" = "Diphtheria Immunization",
                          "Schooling" = "Schooling",
                          "TotalExpenditure" = "Health Expenditure",
                          "pct_doctor" = "Number of Medical Doctors per 10000",
                          "pct_nursing" = "Number of Nurses per 10000")) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| echo: false
#| label: tbl-modelresults
#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.53 years, while in developing countries, it is slightly lower at 71.68 years. This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries. These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and advanced medical technology, contributing to longer life expectancies. "
#| eval: true
#| warning: false
#| message: false
